---
title: "UKBBcleanR: Prepare electronic medical record data from the UK Biobank for time-to-event analyses"
author: 'Derek W. Brown (GitHub: @derekbrown12)'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{UKBBcleanR: Prepare electronic medical record data from the UK Biobank for time-to-event analyses}
  %\VignetteEngine{R.rsp::asis}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = FALSE, fig.show = "hold")
```

## Description

The *UKBBcleanR* package contains an `R` function that prepares time-to-event data from raw [UK Biobank](https://www.ukbiobank.ac.uk/) electronic medical record data. The prepared data can be used for cancer outcomes but could be modified for other health outcomes. This package is not currently available on [CRAN](https://cran.r-project.org/).

## Installation
The release version of the *UKBBcleanR* package can be installed from the GitHub repository by using the *devtools* package:
```{r eval = FALSE}
install.packages("devtools")
library(devtools)
devtools::install_github("machiela-lab/UKBBcleanR")
```

The *UKBBcleanR* package depends on the *data.table*, *dplyr*, and *stringr* packages and may also need to be installed:
```{r eval = FALSE}
install.packages("data.table")
install.packages("dplyr")
install.packages("stringr")
```

Following installation, attach the *UKBBcleanR* package with:
```{r}
library(UKBBcleanR)
```

## Usage
Here we perform a hypothetical example of data cleaning where Chronic Lymphocytic Leukemia (CLL) is the cancer outcome of interest. 

ICD-10 codes, terms, and text within UK Biobank are used by permission of WHO, from the International Statistical Classification of Diseases and Related Health Problems, Tenth Revision (ICD-10). Vols 1-3. Geneva, World Health Organization, 1992-2016. Coding can be downloaded here as a tab-separated file directly from [UK Biobank](https://biobank.ndph.ox.ac.uk/showcase/coding.cgi?tk=bN2V68SNND912GPV53iY2rumWkNQvSs9255965&id=19). A list of ICD-10 coding is also provided [here](https://github.com/machiela-lab/UKBBcleanR/blob/main/data-raw/ICD10_coding.txt). 

### Available Data
Within the *UKBBcleanR* package, we include simulated UK Biobank sample data containing 30 individuals and 490 variables. A detailed list of required variables is provided in the [README_required_variables.txt](https://github.com/machiela-lab/UKBBcleanR/blob/main/data-raw/README_required_variables.txt) file. The raw UK Biobank data can be specified in two ways:

* First option (as demonstrated below). Data can be loaded directly into the `tte` function. The user needs to create a data frame containing all specified variables. Example data is provided in the [`data`](https://github.com/machiela-lab/UKBBcleanR/tree/main/data) folder.

* Second option. The user creates individual `rds` files that contain the variables specified in the [README_required_variables.txt](https://github.com/machiela-lab/UKBBcleanR/blob/main/data-raw/README_required_variables.txt) file. The user needs to specify the working directory using `setwd()` outside the function where these files are stored. Example data is provided in the [`extdata`](https://github.com/machiela-lab/UKBBcleanR/tree/main/inst/extdata) folder. The files are required to be named as follows: 
    + date_enroll.rds
    + cancer_reg.rds
    + inpatient_data.rds
    + self_report_cancer.rds
    + death_reg.rds
    + attrition.rds

```{r}
# Create data frame containing sample data
test_data <- as.data.frame(UKBBcleanR::combined_data)
```

### Set function parameters
For this data cleaning example, we will use CLL (ICD-10 code "C911") as the outcome of interest. We will also identify other non-CLL cancers in the data to control in our analyses. 
```{r}
cancer_outcome <- c("C911") # ICD-10 code for CLL
prevalent_cancers <- c("D37", "D38", "D39", "D40", "D41", "D42", "D43", "D44", "D45", "D46", "D47", "D48") # Prevalent ICD-10 codes to identify - can be blank ex. "c()"
incident_cancers <- c("C900") # Incident ICD-10 codes to identify - can be blank ex. "c()"
```

### Time-to-event data creation
Data creation is performed using `tte`. Within this example, we will not remove anyone identified with prevalent cancer and the time of enrollment. 
```{r}
tte_output_1 <- tte(combined_data = test_data, 
                    cancer_of_interest_ICD10 = cancer_outcome, 
                    prevalent_cancer_list = prevalent_cancers, # Identify specified prevalent cancers
                    prevalent_C_cancers = TRUE, # Include all ICD-10 codes that begin with 'C' to the 'prevalent_cancer_list'
                    incident_cancer_list = incident_cancers, # Identify specified incident cancers 
                    remove_prevalent_cancer = FALSE, # Do not exclude individuals with prevalent cancer(s) or disease(s) 
                    remove_self_reported_cancer = FALSE) # Do not exclude individuals with self reported cancer
```

The generated time-to-event data is output as an object. In the example above, the object is named `tte_output_1`. There are two sets of time-to-event endpoints generated:
* The first ignores user-specified incident cancer (i.e., `incident_cancer_list`). These variables are named `case_control_cancer_ignore` and `censor_days_cancer_ignore` 
* The second controls for user-specified incident cancer (i.e., `incident_cancer_list`). These variables are named `case_control_cancer_control` and `censor_days_cancer_control` 

```{r}
dim(tte_output_1)

table(tte_output_1$case_control_cancer_ignore)
summary(as.numeric(tte_output_1$censor_days_cancer_ignore))

table(tte_output_1$case_control_cancer_control)
summary(as.numeric(tte_output_1$censor_days_cancer_control))
```

From the above data cleaning, there are `30` individuals in the data set with `6` individuals identified as having CLL via `case_control_cancer_ignore` and `1` individual with CLL who also had incident multiple myeloma (ICD-10 code "C900") before their CLL diagnosis via `case_control_cancer_control`. 

The above variables can be used directly in any time-to-event analyses, but there are individuals with CLL identified before enrollment (with negative follow-up time). To remove these individuals and other prevalent cancers, we can rerun the example removing anyone identified to have prevalent cancer and the time of enrollment.
```{r}
tte_output_2 <- tte(combined_data = test_data, 
                    cancer_of_interest_ICD10 = cancer_outcome, 
                    prevalent_cancer_list = prevalent_cancers, # Identify specified prevalent cancers
                    prevalent_C_cancers = TRUE, # Include all ICD-10 codes that begin with 'C' to the 'prevalent_cancer_list'
                    incident_cancer_list = incident_cancers, # Identify specified incident cancers 
                    remove_prevalent_cancer = TRUE, # Exclude individuals with prevalent cancer(s) or disease(s) 
                    remove_self_reported_cancer = TRUE) # Exclude individuals with self reported cancer
``` 

Again, the generated time-to-event data is output as an object. In the example above, the object is named `tte_output_2`.  
```{r}
dim(tte_output_2)

table(tte_output_2$case_control_cancer_ignore)
summary(as.numeric(tte_output_2$censor_days_cancer_ignore))

table(tte_output_2$case_control_cancer_control)
summary(as.numeric(tte_output_2$censor_days_cancer_control))
```

From the above data cleaning, there are `21` individuals in the data set with `3` individuals identified as having CLL via `case_control_cancer_ignore` and `1` individual with CLL who also had incident multiple myeloma (ICD-10 code "C900") before their CLL diagnosis via `case_control_cancer_control`.

The individual with prevalent CLL has now been removed in the `tte_output_2` analysis. The above variables can be used directly in any time-to-event analysis. 

### System setup
The system that rendered this vignette:

```{r system}
sessionInfo()
```
